Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-03-27 manuscript/paper.tex
Good version of Applications section

The goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges 
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale 
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader and the wider 
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web-browser
for display, and dynamically updates as the user interatively navigates the ARG.
This approach allows us to summarise very large ARGs interactively;
Fig~\ref{fig:Supplementary_Figure_4}, for example, shows a screenshot of 
\texttt{tsbrowse} summarising almost 1 million mutations along the SARS-Cov-2
genome.

@@

We demonstrate a further application of tsbrowse in validating inferences
(Supplementary Figures \ref{fig:Supplementary_Figure_2} and
\ref{fig:Supplementary_Figure_3}). Supplementary Figure
\ref{fig:Supplementary_Figure_2} shows how mutation patterns in a 2D plot of
genomic position and mutation age are helpful in spotting poorly inferred
regions. Excluding regions with low site density from the variant data improves
the quality of inference (Panel B). Supplementary Figure
\ref{fig:Supplementary_Figure_3} shows the genomic spans of ancestral nodes in
an ARG inferred with tsinfer. Population genetics theory predicts that IBD
(Identical By Descent) segments break down over time, resulting in shorter
ancestral segments further back in the history of the samples. At time,
\textit{y \textgreater{0.8}}, the plot reveals a pattern of node spans contrary
to this expectation. It confirms a known algorithmic artefact due to which
disproportionately long old ancestors are generated.

Supplementary Fig. \ref{fig:Supplementary_Figure_4}, a screenshot of the
Mutations page for SARS-CoV-2 ARGs \citep{zhan2023towards} demonstrates
tsbrowse's ability to handle diverse data sources. This plot represents a total
of \textcolor{red}{X million} mutations from an inference of
\textcolor{red}{1.27 million} sequences. After executing the
\texttt{preprocess} step, the page loads in \textcolor{red}{X} seconds on a
system with \textcolor{red}{Y GB RAM}, showing that it is possible to visualise
large-scale datasets in a performant way. Panel B demonstrates a useful
interactive feature; additional information about a mutation event (for
example, the mutational frequency in predefined populations, or the number of
sample nodes carrying the mutation) is displayed on mouse-over of individual
data points.

>>>>

The goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges 
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale 
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader and the wider 
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web-browser
for display, and dynamically updates as the user interatively navigates the ARG.
This approach allows us to summarise very large ARGs interactively;
Fig~\ref{fig:Supplementary_Figure_4}, for example, shows a screenshot of 
\texttt{tsbrowse} summarising almost 1 million mutations along the SARS-CoV-2
genome.

@@

Interest in ARG-based methods is burgeoning, but the methods are new 
and practical guidance on applying inferences to real data is lacking.
Data filtering is essential, and the effects of the 
choices that must be made along any bioinformatics pipeline on the final
ARG are hard to predict and quantify. \texttt{Tsbrowse} was primarily
developed as a way to quickly visualise the effects of such filtering
choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensible
element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_2} 
shows [a before and after view for 1000G data, where tsbrowse helped 
diagnose and resolve an opaque data problem].
These interactive visualisations have also helped diagnose issues with
the \texttt{tsinfer} inference algorithm. 
Fig~\ref{fig:Supplementary_Figure_3} shows the genomic spans of 
ancestral nodes in an ARG inferred with tsinfer, demonstrating a clear 
excess of long haplotypes in the very ancient past. These insights 
have helped guide development and 
may lead to significant improvements in performance.

% . Population genetics theory predicts that IBD
% (Identical By Descent) segments break down over time, resulting in shorter
% ancestral segments further back in the history of the samples. At time,
% \textit{y \textgreater{0.8}}, the plot reveals a pattern of node spans contrary
% to this expectation. It confirms a known algorithmic artefact due to which
% disproportionately long old ancestors are generated.

% We demonstrate a further application of tsbrowse in validating inferences
% (Supplementary Figures 
% \ref{fig:Supplementary_Figure_3}). Supplementary Figure
% \ref{fig:Supplementary_Figure_2} shows how mutation patterns in a 2D plot of
% genomic position and mutation age are helpful in spotting poorly inferred
% regions. Excluding regions with low site density from the variant data improves
% the quality of inference (Panel B). 
% Supplementary Figure
% \ref{fig:Supplementary_Figure_3} shows the genomic spans of ancestral nodes in
% an ARG inferred with tsinfer. Population genetics theory predicts that IBD
% (Identical By Descent) segments break down over time, resulting in shorter
% ancestral segments further back in the history of the samples. At time,
% \textit{y \textgreater{0.8}}, the plot reveals a pattern of node spans contrary
% to this expectation. It confirms a known algorithmic artefact due to which
% disproportionately long old ancestors are generated.

% Supplementary Fig. \ref{fig:Supplementary_Figure_4}, a screenshot of the
% Mutations page for SARS-CoV-2 ARGs \citep{zhan2023towards} demonstrates
% tsbrowse's ability to handle diverse data sources. This plot represents a total
% of \textcolor{red}{X million} mutations from an inference of
% \textcolor{red}{1.27 million} sequences. After executing the
% \texttt{preprocess} step, the page loads in \textcolor{red}{X} seconds on a
% system with \textcolor{red}{Y GB RAM}, showing that it is possible to visualise
% large-scale datasets in a performant way. Panel B demonstrates a useful
% interactive feature; additional information about a mutation event (for
% example, the mutational frequency in predefined populations, or the number of
% sample nodes carrying the mutation) is displayed on mouse-over of individual
% data points.
